Leadership across cultures and contexts: an exploratory analysis

### 2020-06-22 leadership_across_cultures_contexts.Rmd
Removed all "see" before Fig references in Discussion section only.

We found strong evidence that several dimensions of leadership were universal across cultures and across contexts within cultures. Leaders are seen as high status, knowledgeable or intelligent, and experienced or accomplished in about 80% or more of cultures, and function to resolve conflict, organize cooperation, and provide counsel or direction in over 70% of cultures (see Figure \@ref(fig:varsupport)). Leaders benefited materially, reproductively, or socially in 50% or more of cultures (see Figure \@ref(fig:varsupportcb)).

@@

The leadership features we derived by clustering leadership dimensions were represented across most cultures. Leaders performed three broad functions: providing prosocial goods and services, strategizing, and organizing collective actions. Regarding the traits characteristic of leadership, leaders possessed some combination of individual capacities, many of which are necessarily prosocial, conforming to particular cultural norms, and receiving various benefits in exchange for their functions. See Figure \@ref(fig:featuressupportplot).

@@

Most potentially universal dimensions discovered here appear in a model developed by @hagen_leadership_2019, in which knowledgeable and intelligent individuals provide computational (cognitive) services to group members in exchange for a variety of fitness benefits, and these services often involve conflict resolution and other social functions [see also @garfield_evolutionary_2019]. The computational services model emphasizes mating benefits, however, which appear in `r matingbenefit`% of cultures, below our arbitrary 60% threshold for "universality" (see Figure \@ref(fig:varsupportcb)). But there was evidence for the *Social, reproductive, and materials success* feature, which included leaders in polygynous unions, having a high quality spouse, and having many children, in more cultures (`r reprosuccess`%) than any other feature (see Figure \@ref(fig:featuressupportplot)). Interestingly, the computational services model highlights women's family leadership role, and the estimated marginal mean for the *Social, reproductive, and materials success* feature was higher for females leaders than male leaders (see Figure S\@ref(fig:heatmapfeaturesex)), with the caveat that there were few text records on female leaders in our sample.

@@

Coercive authority is a dimension of leadership that is common in the ethnographic record and for which evidence does not vary by any of our four factors, yet did not quite make the 60% cutoff to be included on our list of candidate universals. This dimension features in several evolutionary models of leadership involving dominance hierarchies [@barkow_darwin_1989; @tiger_imperial_1971; @henrich_evolution_2001; @cheng_dominance_2019; @van_vugt_dual_2019] and the role of punishment in promoting cooperation [e.g., @marlowe_more_2008; @henrich_markets_2010; @henrich_costly_2006; @boyd_punishment_1992]. Indeed, we found that coercive authority among leaders was associated with enforcing punishment, aggressiveness, being feared, and also with controlling economics, which evokes "resource holding power", a concept from the behavioral ecology of dominance hierarchies. See Figure \@ref(fig:elasticcoercive). This passage describing the authority of Trobriand chiefs is illustrative [@oj29-007, pp. 64-65]:

@@

There was substantial cross-cultural variation in evidence for most dimensions of leadership: of our `r length(all_study_vars)` dimensions of leadership, `r sum(df_compare2$Estimate < 0.5)` had evidence in less than 50% of our `r nrow(leader_cult)` cultures. See Figures \@ref(fig:varsupport) and \@ref(fig:varsupportcb). There was *systematic* variation in `r vary_n` leadership dimensions by group context, subsistence strategy, continental region, and/or leader sex, factors for which we had complete data for all text records.

These four factors were confounded. Specifically, evidence for leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013; @venkataraman_hunter-gatherer_2017]. Evidence for supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories. See Figures \@ref(fig:grouptype) and S\@ref(fig:regionsubsistence). These patterns support perspectives suggesting residential mobility and increased costs of dispersal, subsistence intensification, and demographic transitions are associated with increasing complexity of sociopolitical organization among nonindustrial populations [@johnson_evolution_1987; @steward_basin-plateau_1938; @powers_evolutionary_2014; @kantner_identifying_2010]. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese by @az02-001[p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in *Coding example* section of SI). One interpretation of the negative relationship between evidence for political leadership at the community level or above and leadership in kin groups or residential groups is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes.

@@

The clearest pattern to emerge from analyses of systematic variation is that, not too surprisingly, many dimensions of leadership are context-specific: aggressiveness and bravery were associated with military leadership, supernatural qualities and ritual functions were associated with religious leadership, organizing cooperation and being experience or accomplished were associated with economic-group leadership, and providing counsel and older age were associated with kin-group leadership. See Figure \@ref(fig:heatmapgroups). (Six leadership dimensions varied by continental region; see Figure \@ref(fig:heatmapregion). We do not have clear theoretical interpretations of this variation, so we treat region simply as a control variable.) These results underscore recent trends in evolutionary approaches to leadership that emphasize the context-dependency of leader emergence and evolutionary selection pressures [@smith_leadership_2020; @smith_leadership_2016; @hagen_leadership_2019].

@@

Leaders were widely reported to receive material benefits, social status, and mating benefits but loss of material resources and social status were also widely reported, suggesting leadership to be a high-risk, high-return strategy (see Figure \@ref(fig:varsupportcb)). Nevertheless, for most dimensions of benefits and costs, there was more evidence of benefits than costs (Figure \@ref(fig:benefitcostratio)), and the same was true averaging across all benefit/cost dimensions ($OR=`r signif(leader_bc_ratio$benefit_cost_OR, 2)`$). Additionally, several positive leader qualities indicating social, reproductive, and material success clustered together (bottom cluster Figure \@ref(fig:cluster)A). These results are consistent with much theoretical literature that links leader benefits to their capacities to provide prosocial benefits (e.g., resolving conflicts, organizing collective actions, and being generous) and/or impose social costs (e.g., punishing and aggressiveness) [e.g., @von_rueden_mens_2016; @bowles_emergence_2010; @glowacki_role_2013; @hooper_theory_2010; @gavrilets_solution_2014; @price_evolution_2014; @smith_leadership_2016].

@@

For both leaders and followers, individuals stood to lose what they were able to gain. See Figure \@ref(fig:varsupportcb). @nm09-056 [p. 183] describes Laura Cornelius Kellogg (also known as Minnie and Wynnogene, 1880-1947), a land-claims activist, writer, and perhaps the most famous female leader among the Iroquois, whose life exemplified the risks to both leaders and followers:

@@

Only `r sum(all_data$demo_sex == 'female')` of the `r nrow(all_data)` text records discussed female leaders exclusively, and another `r sum(all_data$demo_sex == 'both')` explicitly discussed both female and male leaders. Leader sex was only associated with variation in seven of our `r length(all_study_vars)` dimensions of leadership (see Figure S\@ref(fig:heatmapsex)), and one feature variable (see Figure S\@ref(fig:heatmapfeaturesex)), indicating that most dimensions of leadership probably characterize leaders of both sexes.

@@

Our results (Figures \@ref(fig:varsupport) and \@ref(fig:varsupportcb)) support the importance of high social status for leader emergence [@cheng_why_2020]. Nevertheless, over two thirds of our text records on leadership did not mention status, and it was not one of our candidate universal dimensions of leadership, with more evidence for high status from South American cultures (see Figure S\@ref(fig:heatmapregion)).

The words and dimensions of leadership predictive of high status leaders (see Figure \@ref(fig:highstatusplot)) suggest the phenotype of such leaders resembles what many anthropologists refer to as "Big men" [@garfield_review_2019]: wealthy, accomplished, older men who favor members of their kin group, and are highly regarded for their decision-making abilities, knowledge, and intelligence. High status leaders are able to direct benefits to their close relatives, likely due, in part, to their greater material and social wealth, which implicates leadership dynamics as a driver of the evolution and emergence of inequality [c.f., @sanday_female_1981; @shenk_status_2016; @price_social_2010; @borgerhoff_mulder_pastoralism_2010; @mattison_evolution_2016; @bowles_emergence_2010]. Our results did not strongly implicate high status leaders with particular functions, with the exception that punishment was modestly negatively associated with evidence for high status. The word "economic" was a negative predictor of high status, hinting that organizational roles might not be high status.

@smith_leadership_2020 speculate that, "leadership may...be more strongly correlated with high status in large, complex organizations, such as in corporations and governments (p. 2)." Evidence for high status leaders, however, did not substantially vary by group structure, which included levels of social organization ranging from sub-residential to supracommunity (see Figure S\@ref(fig:groupstatus)). This null result is evidence against the @smith_leadership_2020 hypothesis.

@@

Shamanism appears to be a distinct form of leadership that combines a strategy of inducing fear, similar to the dominance strategy, but is based on knowledge and expertise, similar to the prestige strategy. The *Competencies* feature included supernatural and feared dimensions (Figure \@ref(fig:cluster)). The MST-kNN analysis similarly clustered leaders with ritual and medicinal functions, which rely on special knowledge, with those who had supernatural, feared or killer qualities (see Figure S\@ref(fig:mstknn)). Our *Shamanism* variable (which was not one of our original leadership dimensions) was predicted by words that evoke both dominance and prestige, such as *supernatural*, *power*, *fear*, and *knowledge* (Figure \@ref(fig:shamanism)A), and by four other leadership dimensions that also overlap with the dominance and prestige constructs: *Feared*, *Medicinal functions*, *Ritual functions*, and *Experience/accomplished*.

>>>>

We found strong evidence that several dimensions of leadership were universal across cultures and across contexts within cultures. Leaders are seen as high status, knowledgeable or intelligent, and experienced or accomplished in about 80% or more of cultures, and function to resolve conflict, organize cooperation, and provide counsel or direction in over 70% of cultures (Figure \@ref(fig:varsupport)). Leaders benefited materially, reproductively, or socially in 50% or more of cultures (Figure \@ref(fig:varsupportcb)).

@@

The leadership features we derived by clustering leadership dimensions were represented across most cultures. Leaders performed three broad functions: providing prosocial goods and services, strategizing, and organizing collective actions. Regarding the traits characteristic of leadership, leaders possessed some combination of individual capacities, many of which are necessarily prosocial, conforming to particular cultural norms, and receiving various benefits in exchange for their functions (Figure \@ref(fig:featuressupportplot)).

@@

Most potentially universal dimensions discovered here appear in a model developed by @hagen_leadership_2019, in which knowledgeable and intelligent individuals provide computational (cognitive) services to group members in exchange for a variety of fitness benefits, and these services often involve conflict resolution and other social functions [see also @garfield_evolutionary_2019]. The computational services model emphasizes mating benefits, however, which appear in `r matingbenefit`% of cultures, below our arbitrary 60% threshold for "universality" (Figure \@ref(fig:varsupportcb)). But there was evidence for the *Social, reproductive, and materials success* feature, which included leaders in polygynous unions, having a high quality spouse, and having many children, in more cultures (`r reprosuccess`%) than any other feature (Figure \@ref(fig:featuressupportplot)). Interestingly, the computational services model highlights women's family leadership role, and the estimated marginal mean for the *Social, reproductive, and materials success* feature was higher for females leaders than male leaders (Figure S\@ref(fig:heatmapfeaturesex)), with the caveat that there were few text records on female leaders in our sample.

@@

Coercive authority is a dimension of leadership that is common in the ethnographic record and for which evidence does not vary by any of our four factors, yet did not quite make the 60% cutoff to be included on our list of candidate universals. This dimension features in several evolutionary models of leadership involving dominance hierarchies [@barkow_darwin_1989; @tiger_imperial_1971; @henrich_evolution_2001; @cheng_dominance_2019; @van_vugt_dual_2019] and the role of punishment in promoting cooperation [e.g., @marlowe_more_2008; @henrich_markets_2010; @henrich_costly_2006; @boyd_punishment_1992]. Indeed, we found that coercive authority among leaders was associated with enforcing punishment, aggressiveness, being feared, and also with controlling economics, which evokes "resource holding power", a concept from the behavioral ecology of dominance hierarchies (Figure \@ref(fig:elasticcoercive)). This passage describing the authority of Trobriand chiefs is illustrative [@oj29-007, pp. 64-65]:

@@

There was substantial cross-cultural variation in evidence for most dimensions of leadership: of our `r length(all_study_vars)` dimensions of leadership, `r sum(df_compare2$Estimate < 0.5)` had evidence in less than 50% of our `r nrow(leader_cult)` cultures (Figures \@ref(fig:varsupport) and \@ref(fig:varsupportcb)). There was *systematic* variation in `r vary_n` leadership dimensions by group context, subsistence strategy, continental region, and/or leader sex, factors for which we had complete data for all text records.

These four factors were confounded. Specifically, evidence for leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013; @venkataraman_hunter-gatherer_2017]. Evidence for supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories (Figures \@ref(fig:grouptype) and S\@ref(fig:regionsubsistence)). These patterns support perspectives suggesting residential mobility and increased costs of dispersal, subsistence intensification, and demographic transitions are associated with increasing complexity of sociopolitical organization among nonindustrial populations [@johnson_evolution_1987; @steward_basin-plateau_1938; @powers_evolutionary_2014; @kantner_identifying_2010]. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese by @az02-001[p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in *Coding example* section of SI). One interpretation of the negative relationship between evidence for political leadership at the community level or above and leadership in kin groups or residential groups is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes.

@@

The clearest pattern to emerge from analyses of systematic variation is that, not too surprisingly, many dimensions of leadership are context-specific: aggressiveness and bravery were associated with military leadership, supernatural qualities and ritual functions were associated with religious leadership, organizing cooperation and being experience or accomplished were associated with economic-group leadership, and providing counsel and older age were associated with kin-group leadership (Figure \@ref(fig:heatmapgroups)). (Six leadership dimensions varied by continental region; Figure \@ref(fig:heatmapregion)). We do not have clear theoretical interpretations of this variation, so we treat region simply as a control variable.) These results underscore recent trends in evolutionary approaches to leadership that emphasize the context-dependency of leader emergence and evolutionary selection pressures [@smith_leadership_2020; @smith_leadership_2016; @hagen_leadership_2019].

@@

Leaders were widely reported to receive material benefits, social status, and mating benefits but loss of material resources and social status were also widely reported, suggesting leadership to be a high-risk, high-return strategy (Figure \@ref(fig:varsupportcb)). Nevertheless, for most dimensions of benefits and costs, there was more evidence of benefits than costs (Figure \@ref(fig:benefitcostratio)), and the same was true averaging across all benefit/cost dimensions ($OR=`r signif(leader_bc_ratio$benefit_cost_OR, 2)`$). Additionally, several positive leader qualities indicating social, reproductive, and material success clustered together (bottom cluster Figure \@ref(fig:cluster)A). These results are consistent with much theoretical literature that links leader benefits to their capacities to provide prosocial benefits (e.g., resolving conflicts, organizing collective actions, and being generous) and/or impose social costs (e.g., punishing and aggressiveness) [e.g., @von_rueden_mens_2016; @bowles_emergence_2010; @glowacki_role_2013; @hooper_theory_2010; @gavrilets_solution_2014; @price_evolution_2014; @smith_leadership_2016].

@@

For both leaders and followers, individuals stood to lose what they were able to gain (Figure \@ref(fig:varsupportcb)). @nm09-056 [p. 183] describes Laura Cornelius Kellogg (also known as Minnie and Wynnogene, 1880-1947), a land-claims activist, writer, and perhaps the most famous female leader among the Iroquois, whose life exemplified the risks to both leaders and followers:

@@

Only `r sum(all_data$demo_sex == 'female')` of the `r nrow(all_data)` text records discussed female leaders exclusively, and another `r sum(all_data$demo_sex == 'both')` explicitly discussed both female and male leaders. Leader sex was only associated with variation in seven of our `r length(all_study_vars)` dimensions of leadership (Figure S\@ref(fig:heatmapsex)), and one feature variable (Figure S\@ref(fig:heatmapfeaturesex)), indicating that most dimensions of leadership probably characterize leaders of both sexes.

@@

Our results (Figures \@ref(fig:varsupport) and \@ref(fig:varsupportcb)) support the importance of high social status for leader emergence [@cheng_why_2020]. Nevertheless, over two thirds of our text records on leadership did not mention status, and it was not one of our candidate universal dimensions of leadership, with more evidence for high status from South American cultures (Figure S\@ref(fig:heatmapregion)).

The words and dimensions of leadership predictive of high status leaders (Figure \@ref(fig:highstatusplot)) suggest the phenotype of such leaders resembles what many anthropologists refer to as "Big men" [@garfield_review_2019]: wealthy, accomplished, older men who favor members of their kin group, and are highly regarded for their decision-making abilities, knowledge, and intelligence. High status leaders are able to direct benefits to their close relatives, likely due, in part, to their greater material and social wealth, which implicates leadership dynamics as a driver of the evolution and emergence of inequality [c.f., @sanday_female_1981; @shenk_status_2016; @price_social_2010; @borgerhoff_mulder_pastoralism_2010; @mattison_evolution_2016; @bowles_emergence_2010]. Our results did not strongly implicate high status leaders with particular functions, with the exception that punishment was modestly negatively associated with evidence for high status. The word "economic" was a negative predictor of high status, hinting that organizational roles might not be high status.

@smith_leadership_2020 speculate that, "leadership may...be more strongly correlated with high status in large, complex organizations, such as in corporations and governments (p. 2)." Evidence for high status leaders, however, did not substantially vary by group structure, which included levels of social organization ranging from sub-residential to supracommunity (Figure S\@ref(fig:groupstatus)). This null result is evidence against the @smith_leadership_2020 hypothesis.

@@

Shamanism appears to be a distinct form of leadership that combines a strategy of inducing fear, similar to the dominance strategy, but is based on knowledge and expertise, similar to the prestige strategy. The *Competencies* feature included supernatural and feared dimensions (Figure \@ref(fig:cluster)). The MST-kNN analysis similarly clustered leaders with ritual and medicinal functions, which rely on special knowledge, with those who had supernatural, feared or killer qualities (Figure S\@ref(fig:mstknn)). Our *Shamanism* variable (which was not one of our original leadership dimensions) was predicted by words that evoke both dominance and prestige, such as *supernatural*, *power*, *fear*, and *knowledge* (Figure \@ref(fig:shamanism)A), and by four other leadership dimensions that also overlap with the dominance and prestige constructs: *Feared*, *Medicinal functions*, *Ritual functions*, and *Experience/accomplished*.
